---
title: "Exercise 5: Introduction to Spatial Omics Data Analysis"
author: "James Khaznadar"
date: "`r Sys.Date()`"
output:
  workflowr::wflow_html:
    self_contained: true
---

> Full rendered version with plots: https://jkhazn.github.io/BIO334_Robinson/khaznadar_james_exercise5.html

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(SpatialExperiment)
library(STexampleData)
library(ggplot2)
library(dplyr)
library(Voyager)
library(scrapper)
library(scater)
library(BayesSpace)
library(Banksy)
library(tidyr)
library(pals)
library(scran)
```

## Load Data

```{r load-data}
spe <- Visium_humanDLPFC()
spe
```

## 2.1 Exploring the Metadata

How many genes and spots? How many genes have non-zero expression? How many spots are in tissue? Show distribution of cells per spot. Plot ground truth spatial domains.

There are 33538 genes and 4992 spots.

```{r q1-dims}
dim(spe)
sum(rowSums(counts(spe)) > 0)
```


Of the 4992 spots, 3639 are in tissue.
```{r q2-in-tissue}
table(spe$in_tissue)
```

```{r q3-cells-per-spot}

hist(spe$cell_count, breaks = 10)
ggplot(data.frame(cell_count = spe$cell_count), aes(x = cell_count)) +
  geom_density() +
  xlab('Count') +
  ylab('Frequency')
```

```{r q4-ground-truth}
data.frame(colData(spe), spatialCoords(spe)) |>
  ggplot(aes(x = pxl_col_in_fullres, y = pxl_row_in_fullres, color = ground_truth)) +
  geom_point() +
  coord_equal() +
  scale_y_reverse() +
  scale_color_discrete(na.value = "white") +
  theme_void()

```

## 2.2 Filtering and QC

Explore QC metrics and apply reasonable filters.

```{r qc}
spe <- spe[, spe$in_tissue == 1]

is_mito <- grepl("(^MT-)|(^mt-)", rowData(spe)$gene_name)
table(is_mito)

spe <- quickRnaQc.se(spe, subsets = list(MT = is_mito))

plotColData(spe, x = "detected", y = "subset.proportion.MT",
            colour_by = "keep") + scale_x_sqrt()

data.frame(colData(spe), spatialCoords(spe)) |>
  ggplot(aes(x = pxl_col_in_fullres, y = pxl_row_in_fullres, color = keep)) +
  geom_point() +
  coord_equal() +
  scale_y_reverse() +
  theme_void()
```

```{r filter}
spe <- spe[, colData(spe)$keep]
spe <- spe[rowSums(counts(spe)) > 0, ]
dim(spe)
```

## 2.3 Normalization and Dimension Reduction

Normalize, find HVGs, PCA. What layers of DLPFC are transcriptionally distinct?

```{r normalize}
spe <- normalizeRnaCounts.se(spe)
```

```{r hvg-pca}
spe <- chooseRnaHvgs.se(spe, top = 2000)
plot(rowData(spe)$means, rowData(spe)$variances, col=factor(rowData(spe)$hvg))
is.hvg <- rowData(spe)$hvg    

# Get the gene names:
hvg.names <- rownames(spe)[is.hvg]
length(hvg.names)
```

Plot the first, second and third PC in spatial context.

```{r pcs-spatial}
spe <- runPCA(spe, subset_row = hvg.names, ncomponents = 50)
df <- data.frame(colData(spe), reducedDim(spe, "PCA"), spatialCoords(spe))
#rename the colnames of the dataframe
colnames(df) <- gsub("^X(\\d+)$", "PC\\1", colnames(df))
df |>
  dplyr::select(c(PC1:PC3, pxl_col_in_fullres, pxl_row_in_fullres)) |>
  pivot_longer(PC1:PC3) |>
  ggplot(aes(x = pxl_col_in_fullres, y = pxl_row_in_fullres, color = value)) +
  geom_point(size = 0.5) +
  facet_wrap( ~ name) +
  coord_equal() + # to make x and y coordinates have the same spacing
  scale_color_gradientn(colors = pals::parula()) + # custom color scale
  scale_y_reverse() +
  theme_void()
```

Use Banksy for spatially-aware dimension reduction. Does it improve things?
Slightly, the white matter is now clearly clustered together, with there being some other higher level structures visible which is closer to the groud truth.

```{r banksy}
set.seed(123)
# select number of neighbors for the shared nearest-neighbor graph
num.neighbors <- 15
# run the graph clustering
spe <- clusterGraph.se(spe, num.neighbors=num.neighbors)
data.frame(colData(spe), spatialCoords(spe)) |>
  ggplot(aes(x = pxl_col_in_fullres, y = pxl_row_in_fullres, color = clusters)) +
  geom_point() +
  coord_equal() + # to make x and y coordinates have the same spacing
  theme_void() +
  scale_color_manual(values = pals::brewer.set1(length(unique(spe$clusters)))) +
  scale_y_reverse()
```

## 2.4 Clustering into Spatial Domains

Apply spatially-aware clustering, compare to ground truth with ARI, plot side by side.

```{r cluster}
set.seed(123)
spe <- computeBanksy(spe, assay_name = "logcounts")
spe <- runBanksyPCA(spe)
spe <- clusterBanksy(spe)
data.frame(colData(spe), spatialCoords(spe)) |>
  ggplot(aes(x = pxl_col_in_fullres, y = pxl_row_in_fullres, color = clust_M0_lam0.2_k50_res1)) +
  geom_point() +
  coord_equal() +
  theme_void() +
  scale_color_manual(values = pals::brewer.set2(length(unique(spe$clust_M0_lam0.2_k50_res1)))) +
  scale_y_reverse()
```

```{r ari}
library(mclust)
idx <- !is.na(spe$ground_truth)
adjustedRandIndex(spe$ground_truth[idx], spe$clust_M0_lam0.2_k50_res1[idx])
```

```{r cluster-vs-truth}
data.frame(colData(spe), spatialCoords(spe)) |>
  ggplot(aes(x = pxl_col_in_fullres, y = pxl_row_in_fullres, color = ground_truth)) +
  geom_point() +
  coord_equal() +
  scale_y_reverse() +
  scale_color_discrete(na.value = "grey80") +
  theme_void() +
  ggtitle("Ground truth")

data.frame(colData(spe), spatialCoords(spe)) |>
  ggplot(aes(x = pxl_col_in_fullres, y = pxl_row_in_fullres, color = clust_M0_lam0.2_k50_res1)) +
  geom_point() +
  coord_equal() +
  scale_y_reverse() +
  theme_void() +
  scale_color_manual(values = pals::brewer.set2(length(unique(spe$clust_M0_lam0.2_k50_res1)))) +
  ggtitle("Banksy clusters")
```

## 2.5 Moran's I

Compute Moran's I for the 3 genes from the paper + 200 random HVGs. Plot and highlight the 3 genes.

```{r morans-i}
paper_genes <- c("MOBP", "PCP4", "SNAP25")
paper_ids <- rownames(spe)[rowData(spe)$gene_name %in% paper_genes]
set.seed(123)
test_genes <- union(paper_ids, sample(hvg.names, 200))

sfe <- toSpatialFeatureExperiment(spe)
colGraph(sfe, "visium") <- findVisiumGraph(sfe, zero.policy = TRUE)
sfe <- runUnivariate(sfe, type = "moran", features = test_genes,
                     colGraphName = "visium", zero.policy = TRUE)

plotSpatialFeature(sfe, paper_ids, ncol = 3, maxcell = 5e4, swap_rownames = "gene_name")
```
