---
title: "Exercise 4: Single-cell RNA-seq"
author: "James Khaznadar"
date: "`r Sys.Date()`"
output:
  workflowr::wflow_html:
    self_contained: true
---

> Full rendered version with plots: https://jkhazn.github.io/BIO334_Robinson/khaznadar_james_exercise4.html

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
suppressPackageStartupMessages({
  library(SingleCellExperiment)
  library(scrapper)
  library(scater)
  library(CellBench)
  library(ExperimentHub)
  library(SingleR)
  library(pheatmap)
  library(ensembldb)
  library(patchwork)
  library(dplyr)
})
nthreads <- 2
```

## Getting Started

How many cells do we have for each technology? How many genes? Where is the cell line information stored?

```{r load-data}
tenx     <- load_sc_data()[["sc_10x"]]
cel_seq  <- load_sc_data()[["sc_celseq"]]
drop_seq <- load_sc_data()[["sc_dropseq"]]
```

```{r explore}
dim(tenx)
dim(cel_seq)
dim(drop_seq)

head(tenx$cell_line)
```

## Merge Datasets

What are potential drawbacks of merging by intersecting genes?

```{r merge}
sce_list <- list(tenx, drop_seq, cel_seq) %>%
  set_names(c("tenx", "drop_seq", "celseq"))

intersect_genes <- Reduce(intersect, lapply(sce_list, rownames))

overlap_cols <- Reduce(intersect,
  lapply(sce_list, function(u) names(colData(u))))

sce_list <- lapply(sce_list, function(sce) {
  sce_sub <- sce[intersect_genes, ]
  colData(sce_sub) <- colData(sce)[, overlap_cols]
  sce_sub
})

sce_com <- do.call(SingleCellExperiment::cbind, sce_list)

sce_com$technology <- rep(names(sce_list), sapply(sce_list, ncol))

colnames(sce_com) <- rownames(colData(sce_com)) <- paste0(sce_com$technology, "_", colnames(sce_com))

rm(sce_list, tenx, drop_seq, cel_seq)
gc()
```

## Combined Dataset

How many cells and genes are left? Which assays are present? Where is the technology info?

```{r explore-combined}
dim(sce_com)
assayNames(sce_com)
head(sce_com$technology)
```

## Pre-processing

```{r normalize}
sce_com <- normalizeRnaCounts.se(sce_com)
```

```{r hvg}
sce_com <- chooseRnaHvgs.se(sce_com, top = 1500, num.threads = nthreads)
top.hvgs <- which(rowData(sce_com)$hvg)
```

```{r pca-umap}
set.seed(00101)
sce_com <- runPca.se(sce_com, features = top.hvgs, num.threads = nthreads)
sce_com <- runUmap.se(sce_com, num.dim = 2, num.threads = nthreads)
```

## Question 1 — Batch Effect

Do you find a batch effect caused by the different technologies? Which technologies are most similar?

```{r q1-batch}
plotUMAP(sce_com, colour_by = "technology")
plotUMAP(sce_com, colour_by = "cell_line")
```

Yes, there is a batch effect. CELseq clusters separately, 10x and Drop-seq are most similar.

## Question 2 — Clustering

Do cells of different cell lines cluster separately?

```{r q2-cluster}
set.seed(00101)
sce_com <- clusterGraph.se(sce_com, num.threads = nthreads, output.name = "cluster")
sce_com$cluster <- factor(sce_com$cluster)

plotUMAP(sce_com, colour_by = "cluster")
plotUMAP(sce_com, colour_by = "cell_line")
table(sce_com$cluster, sce_com$cell_line)
```

Yes, cells cluster by cell line as confirmed by the UMAP and table, but maybe change count of clusters, i would probably use a cluster size of 3 given the three cell lines to tell.

## Question 3 — Marker Genes (optional)

What are the marker genes for each cluster? How many markers do you find? Why do we only find so few markers for some clusters?

```{r q3-markers}

```

## Question 4 — Cell Type Annotation (optional)

Use a reference dataset from ExperimentHub to annotate the cells. Do you see a problem with the annotation?

```{r q4-annotation}

```
