---
title: "Exercise 2: dplyr and tidyr for data frames"
author: "James Khaznadar"
date: "`r Sys.Date()`"
output:
  workflowr::wflow_html:
    self_contained: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
rm(list=ls())
library(readr)
library(dplyr)
library(tidyr)
library(tibble)
```

## Question 1

Get a count matrix for RNA-seq data from the [airway](http://bioconductor.org/packages/release/data/experiment/html/airway.html) data set. In this experiment, four different human airway smooth muscle cell lines were treated with dexamethasone. The package provides read counts per gene for four control and the corresponding treatment samples.

```{r load-airway}
library(airway) ## load the package
data("airway")  ## load the data

```

Rewrite the following code to fill the data into data frames with `magrittr` pipes.

```{r q1-pipes}
md <- airway %>%
  colData() %>%
  as.data.frame()

expr <- airway %>%
  assay() %>%
  as.data.frame() %>%
  rownames_to_column("feature")
```

* Select the first 3 genes of the count matrix and save it as a new data frame.
* Transform the new data frame to long format with 3 columns: `sample`, `gene` and `counts`.
* Add a new column to the data frame with the condition (`dex`) of each sample from `md$dex`.

```{r q1-long-format}
first_three_genes <- head(expr, 3)

longer <- first_three_genes %>%
  pivot_longer(-feature, names_to = "sample", values_to = "counts") %>%
  dplyr::rename(gene = feature) %>%
  mutate(condition = md[sample, "dex"]) %>%
  select(sample, condition, gene, counts)
```
Count how many samples we have for each gene.

```{r q1-count}
gene_count <- longer %>%
  group_by(gene) %>%
  dplyr::count()
```

## Question 2

Dr. Granger is interested in studying the factors controlling the size and carbon storage of shrubs. She has conducted a small preliminary experiment looking at the effect of three different treatments on shrub volume at four different locations. Data from this [Data Carpentry lesson](https://datacarpentry.org/semester-biology/data/shrub-volume-data.csv).

* Read the data from the file `shrub-volume-data.csv`.

```{r load-shrub}
shrub <- read.csv('https://datacarpentry.org/semester-biology/data/shrub-volume-data.csv')
```

* Replace the three dimension columns with a volume column and store the result in a new data frame.

```{r q2-volume}
shrub <- mutate(shrub,volume = length*width*height) %>%
   select(-length, -width, -height)
```

* Join the manipulation from `experiments_table` with the new data frame.

```{r q2-join}
experiments_table <- tibble(experiment = 1:3, manipulation = c("control", "burn", "rainout"))
experiments_table <- shrub %>% full_join(experiments_table, by = "experiment")

```

* Site 4 was partially contaminated with fertilizer — remove it from the data frame.

```{r q2-clean}
experiments_clean <- experiments_table %>%
  filter(site != 4)

```

* Change `"rainout"` to `"radioactive rainout"` in the `manipulation` column.

```{r q2-rename}
experiments_clean <- experiments_clean %>%
  mutate(manipulation = case_match(manipulation, "rainout" ~ "radioactive rainout", .default = manipulation))

```

* Show the mean, minimum and maximum volume for each manipulation, arranged from highest to lowest mean.

```{r q2-summary}
experiment_summary <- experiments_clean %>%
  group_by(manipulation) %>%
  summarise(mean_v = mean(volume, na.rm = TRUE), min_v = min(volume, na.rm = TRUE), max_v = max(volume, na.rm = TRUE)) %>%
  arrange((mean_v))

```

## Question 3

We have gene expression count data from 7 files comparing S2 cells that were untreated (4 samples) versus knocking down the splicing factor pasilla/CG8144 (3 samples) in Drosophila. Each file contains 2 tab-separated columns: gene identifier and count.

* Read in the 7 files, add sample information to each data frame, and bind them into a single data frame using `lapply`.

```{r load-counts}
fs <- list.files(here::here("data", "files_for_ex1"), ".count$", full.names = TRUE)
my_function <- function(filepath) {
  df <- read.table(filepath)
  df$sample <- basename(filepath)
  df
}

counts_df <- lapply(fs, my_function) %>% 
  bind_rows()

```

* Calculate the CPM and count the number of genes with CPM > 1 for each sample.

```{r q3-cpm}
cpm_answer <- counts_df %>%
  group_by(sample) %>%
  mutate(CPM = 1e6 * V2 / sum(V2)) %>%
  summarise(n = sum(CPM > 1))

```

* Transform the long count data frame into a wide count matrix with genes as `rownames()` and samples as `colnames()`. Include only genes in rows 10 to 15.

```{r q3-wide}

thick_df <- counts_df %>%
  pivot_wider(names_from = sample, values_from = V2) %>%
  column_to_rownames("V1") %>%
  dplyr::slice(10:15)


```
